---
title: "Using time series to predict popularity of names."
author: "Christopher C. Jackson"
output: github_document
---

##Introduction:

I chose to analyze the popularity (ranking) of baby girl names based on data provided by the Social Security Administration website during the period between 1980 and 2017. I observed trends and used time series analysis to predict baby girl rankings at future times. I restricted my study to the five most popular names of 1980 (Jennifer, Amanda, Jessica, Melissa, Sarah).

```{r}
namesdata<-read.csv("babydata.csv")
head(namesdata)

```
##Graphing data: Visualizing the evolution of rankings over the 1980-2017 period.

```{r echo=FALSE}
plot(namesdata$Year,namesdata$Jennifer_Rank,type="o",col="red",xlab="Year",ylim=c(1,400),ylab="Rank in Popularity ",pch=15, main="Rank of selected baby girl names")
lines(namesdata$Year,namesdata$Amanda_Rank, type="o", col="blue",pch=16)
lines(namesdata$Year, namesdata$Jessica_Rank, type="o",col="green",pch=17)
lines(namesdata$Year, namesdata$Melissa_Rank, type="o", col="orange",pch=18)
lines(namesdata$Year,namesdata$Sarah_Rank,type="o",col="black",pch=19)
legend("topleft",c("Jennifer","Amanda","Jessica","Melissa","Sarah"),col=c("red","blue","green","orange","black"), pch=15:19)
```

##Analysis:

Based on the graph above, the five names-Jennifer,Amanda,Jessica,Melissa,Sarah-which all ranked between 1 and 5 in popularity in 1980 now range between 62 and 334 in rank today. In other words, every one of those names has significantly decreased in popularity. 

During the 1980-1995 period, the names generally stayed close in rank to 1980 values.

During the 1995-2005 period, the names generally increased in rank (became less popular) more rapidly than during the 1980-1995 period.

After 2005, all names generally increased in rank very rapidly compared to the preceeding periods.


##Time Series: Using R's time-series functions to predict future rankings.

###Part 1: Let us use data between 1980 and 2015 to predict baby girl name rankings in 2016 and 2017; we can then compare those values to the actual rankings. 

####Time series for "Jennifer" during 1980-2015 period:
```{r}
library(forecast)
myseries1<-ts(namesdata$Jennifer_Rank,start=c(1980,1),end=c(2015),frequency=1)
fit<-ets(myseries1,model="AAN")
fit
forecast(fit,2)
plot(forecast(fit,2))
```

####Prediction and comparision of rank of "Jennifer" in 2016 and 2017:

Using time series analysis based on 1980-2015 data, R estimates a rank of 279 (in 2016) and a rank of 308 (in 2017) for Jennifer. According to the Social Security Administration, actual ranks for Jennifer are 273 (in 2016) and 310 (in 2017). 

The real 2016 value of 273 falls inside the 95 percent confidence interval (267,292) generated by R.
The real 2017 value of 310 falls inside the 95 percent confidence interval (286,331) generated by R.

####Time series for "Amanda" during 1980-2015 period:
```{r}
library(forecast)
myseries2<-ts(namesdata$Amanda_Rank,start=c(1980,1),end=c(2015),frequency=1)
fit1<-ets(myseries2,model="AAN")
fit1
forecast(fit1,2)
plot(forecast(fit1,2))
```

####Prediction and comparision of rank of "Amanda" in 2016 and 2017:

Using time series analysis based on 1980-2015 data, R estimates a rank of 318 (in 2016) and a rank of 320 (in 2017) for Amanda. Actual ranks for Amanda are 327 (in 2016) and 334 (in 2017).

The real 2016 value of 327 falls inside the 95 percent confidence interval (306,331) generated by R.
The real 2017 value of 334 falls inside the 95 percent confidence interval (294,349) generated by R.

#####Time series for "Jessica" during 1980-2015 period:
```{r}
library(forecast)
myseries3<-ts(namesdata$Jessica_Rank,start=c(1980,1),end=c(2015),frequency=1)
fit2<-ets(myseries3,model="AAN")
fit2
forecast(fit2,2)
plot(forecast(fit2,2))
```

####Prediction and comparision of rank of "Jessica" in 2016 and 2017:

Using time series analysis based on 1980-2015 data, R estimates a rank of 226 (in 2016) and a rank of 249 (in 2017) for Jessica. Actual ranks for Jessica are 233 (in 2016) and 248 (in 2017). 

The real 2016 rank of 233 falls inside the 95 percent confidence interval (220,234) generated by R.
The real 2017 rank of 248 falls inside the 95 percent confidence interval (238,260) generated by R.

####Time series for "Melissa" during 1980-2015 period:

```{r}
library(forecast)
myseries4<-ts(namesdata$Melissa_Rank,start=c(1980,1),end=c(2015),frequency=1)
fit3<-ets(myseries4,model="AAN")
fit3
forecast(fit3,2)
plot(forecast(fit3,2))
```

####Prediction and comparision of rank of "Melissa" in 2016 and 2017:

Using time series analysis based on 1980-2015 data, R estimates a rank of 262 (in 2016) and a rank of 287 (in 2017) for Melissa. Actual ranks for Melissa are 259 (in 2016) and 273 (in 2017). 

The real 2016 value of 259 falls inside the 95 percent confidence interval (249,277) generated by R.
The real 2017 value of 273 inside the 95 percent confidence interval (270,304) generated by R.

####Time series for "Sarah" during 1980-2015 period:

```{r}
library(forecast)
myseries5<-ts(namesdata$Sarah_Rank,start=c(1980,1),end=c(2015),frequency=1)
fit4<-ets(myseries5,model="AAN")
fit4
forecast(fit4,2)
plot(forecast(fit4,2))
```

####Prediction and comparision of rank of "Sarah" in 2016 and 2017:

Using time series analysis based on 1980-2015 data, R estimates a rank of 63 (in 2016) and a rank of 68 (in 2017) for Sarah. Actual ranks for Sarah are 57 (in 2016) and 62 (in 2017).

The real 2016 value of 57 falls outside the 95 percent confidence interval (60,67) generated by R.
The real 2017 value of 62 falls inside the 95 percent confidence interval (62,75) generated by R.

###Part 2: Let us use data between 1980 and 2017 to predict baby girl rankings in 2018,2019,and 2020.

####Time series for "Jennifer" during 1980-2017 period:

```{r}
library(forecast)
myseries6<-ts(namesdata$Jennifer_Rank,start=c(1980,1),end=c(2017),frequency=1)
fit5<-ets(myseries6,model="AAN")
fit5
forecast(fit5,3)
plot(forecast(fit5,3))
```

####Predictions for rank of "Jennifer" in 2018, 2019, and 2020:

Using time series analysis based on the 1980-2017 data, R estimates a rank of 340 for Jennifer in 2018 with a 95-percent confidence interval of (328,353). R estimates a rank of 371 for Jennifer in 2019 with a 95-percent confidence interval of (349,394). R estimaes a rank of 401 for Jennifer in 2020 with a 95-percent confidence interval (368,435).

####Time series for "Amanda" during 1980-2017 period:

```{r}
library(forecast)
myseries7<-ts(namesdata$Amanda_Rank,start=c(1980,1),end=c(2017),frequency=1)
fit6<-ets(myseries7,model="AAN")
fit6
forecast(fit6,3)
plot(forecast(fit6,3))
```

####Predictions for rank of "Amanda" in 2018, 2019, and 2020:

Using time series analysis based on the 1980-2017 data, R estimates a rank of 341 for Amanda in 2018 with a 95-percent confidence interval of (329,354). R estimates a rank of 348 for Amanda in 2019 with a 95-percent confidence interval of (322,376). R estimaes a rank of 356 for Amanda in 2020 with a 95-percent confidence interval (312,400).

#### Time series for "Jessica" during 1980-2017 period:

```{r}
library(forecast)
myseries8<-ts(namesdata$Jessica_Rank,start=c(1980,1),end=c(2017),frequency=1)
fit7<-ets(myseries8,model="AAN")
fit7
forecast(fit7,3)
plot(forecast(fit7,3))
```

#### Predictions for rank of "Jessica" in 2018, 2019, and 2020:

Using time series analysis based on the 1980-2017 data, R estimates a rank of 272 for Jessica in 2018 with a 95-percent confidence interval of (264,280). R estimates a rank of 293 for Jessica in 2019 with a 95-percent confidence interval of (282,305). R estimaes a rank of 314 for Jessica in 2020 with a 95-percent confidence interval (297,333).

####Time series for "Melissa" during 1980-2017 period:

```{r}
library(forecast)
myseries9<-ts(namesdata$Melissa_Rank,start=c(1980,1),end=c(2017),frequency=1)
fit8<-ets(myseries9,model="AAN")
fit8
forecast(fit8,3)
plot(forecast(fit8,3))
```

#### Predictions for rank of "Melissa" in 2018, 2019, and 2020

Using time series analysis based on the 1980-2017 data, R estimates a rank of 300 for Melissa in 2018 with a 95-percent confidence interval of (287,315). R estimates a rank of 321 for Melissa in 2019 with a 95-percent confidence interval of (305,337). R estimaes a rank of 341 for Melissa in 2020 with a 95-percent confidence interval (322,361).

#### Time series for "Sarah" during 1980-2017 period

```{r}
library(forecast)
myseries10<-ts(namesdata$Sarah_Rank,start=c(1980,1),end=c(2017),frequency=1)
fit9<-ets(myseries10,model="AAN")
fit9
forecast(fit9,3)
plot(forecast(fit9,3))
```

####Predictions for rank of "Sarah" in 2018, 2019, and 2020

Using time series analysis based on the 1980-2017 data, R estimates a rank of 65 for Sarah in 2018 with a 95-percent confidence interval of (62,70). R estimates a rank of 69 for Sarah in 2019 with a 95-percent confidence interval of (63,76). R estimaes a rank of 73 for Sarah in 2020 with a 95-percent confidence interval (64,83).

##Conclusion:

I observed that using time series was pretty accurate to predict rankings for 2016 and 2017 based on 1980-2015 data. In all but one case, actual 2016 and 2017 rankings fell within the 95 percent confidence interval generated by R. Based off the success of the 1980-2015 study, I used 1980-2017 data to make predictions for 2018, 2019, 2020 rankings. Overall, I notice that all five names are increasing in rank through 2020, meaning that all five names are becoming less popular choices for baby girls. I am suprised to see that all of the five most popular names in 1980 have dropped so dramatically in popularity.
